---
title: "College Major Backgrounds and Effects"
author: "Dylan Holliday and Max Collins"
date: "May 7, 2025"
date-modified: now
format: 
  pdf:
    embed-resources: true
    number-sections: false
    toc: false
    fig-align: center
    fig-height: 5
    cap-location: top
    link-external-newwindow: true
execute: 
  echo: false
  warning: false
---

# Research Topic: College Majors

Our Research is focused on finding trends and insights from post-secondary majors: how one's socioeconomic status plays a role in their major, as well as their major's effect on their income after graduation. This is a topic of interest for our team because we are in college and are thus interested in how different majors contain different groups of people and how the choice of a college major can affect one's future career. We will explore these relationships using a series of data visualizations and provide insights on income and diversity trends of recent college graduates. Our underlying purpose is to expose societal trends that may be hindering diversity initiatives in order to become a basis for possible change.

# Research Questions

We plan to conduct an exploratory data analysis from the information given by our data sets. Our first question we will explore is what is the highest earning major category. We will find the said category and perform a two sample t test to show if the highest earning category is significantly different from the 2nd highest. We then want to know about the distribution of the highest earning major category to see if there is any skew or outliers that would affect the mean. We also inquire whether there is a significant difference in college race proportions compared to the national proportions. To do this, we intend to conduct a one sample proportion test comparing primarily the white proportion of college students to the national to expose a possible bias. 

# Provenance of Our Data

Our primary data source was found on Kaggle. Kaggle is a data science focused website that contains many data sets that learners can find and use. The author of this particular data set is Bojan Tunguz. The data is described as being from the American Community Survey 2010-2012 Public Use Microdata Series. The data set contains multiple data frames, but our study will focus upon the recent-grads data, where each case represents a unique major. Each case contains information on majors, earnings post-graduation, employment, employment type, major categories, and the amount of people in each major. Since these data are from a specific survey, they have not been updated and will not. The data, however, still maintains integrity following that all these majors still exist and no abnormal world events would lead us to deny insight from the data. Our analysis here will mainly focus on median salary, major categories (majors grouped within a similar field), unemployment rate, and total amount of graduates.  

Our secondary data source is from the 2022 American Community Survey, from the U.S. Census Bureau. Each case for this data set is also a major, so that is how we intend to join the data frames together. This data contains information on popular majors and the percentage of each race and sex, as well as age and income categories. With all this said, our research with this data set will solely focus on the race proportions and poverty proportions. These data are not to be updated since they are also from a specific survey. Nevertheless, the data provides useful insights on socioeconomic background's role in college majors.

# FAIR and CARE Principles

The data that we used through this project satisfies the FAIR principles of findability, accessibility, interoperability, and reusability. The first dataset we found came from a publicly accessible repository from FiveThirtyEight, a reputable source on data analysis and reporting, and the second dataset we used came from the publicly accessible US Census database. These data sets can be accessed by anyone on the internet as they have been freely provided by the US Census Bureau and FiveThirtyEight. The data was also in compatible formats such as .csv files and .xlsx files, and all languages used throughout our report and the wrangling and visualization process are formal and applicable to the analysis we conducted. And finally, the data we used contains plenty of relevant pieces of economic and socioeconomic data, which, as outlined above, have sufficient provenance. 

Our data also satisfies the CARE principles of collective benefit, authority to control, responsibility, and ethics. These standards are especially relevant to our project as we are working with socioeconomic data that could potentially be sensitive to some people. Our analysis aims to discuss socioeconomic impacts and trends associated with a choice in college major and give people insight into our potential findings on their impact on economic outcomes. The data we used also came from the US Census, which occurs every 10 years, and people willingly participate in. We are trying to use this data to expand people’s understanding of how college major choices relate to income, poverty, and unemployment. And finally, we are not trying to do any harm through this project and are doing this to explore trends that may allow people to make better decisions for their future. 


# Analyses

```{r}
### Read in data and format to appropriate data frames

#1 Open Necessary Packages
library(tidyr)
library(readr)
library(dplyr)
library(rvest)
library(tidyverse)
library(ggplot2)
library(readxl)
library(janitor)
library(knitr)
library(kableExtra)

#2 Read in earnings data from url, diversity data from xlsx path

# Read in income data
recentGrads <- read_csv(
  # Separate string for visibility
  str_c("https://raw.githubusercontent.com/DJH6655",
        "/STAT184-Final-Project/refs/heads/main/recent-grads.csv"))

# Read in secondary data source
sexRaceFreqRaw <- read_excel("SexRaceFrequency.xlsx")


#3 Clean Excel Data, Selecting necessary attributes and renaming them
sexRaceFrequency <- sexRaceFreqRaw %>% 
  slice_head(n=41) %>% 
  slice_tail(n = 37) %>% 
  select(c(1,2,9,10,11,12,13,21,22)) %>% 
  rename(
    Major = 1,
    Total = 2,
    Race_White_Percent = 3, 
    Race_Black_Percent = 4,
    Race_Asian_Percent = 5,
    Race_Hispanic_Percent = 6,
    Race_OtherMulti_Percent = 7,
    Poverty_Percent = 8,
    Not_Poverty_Percent = 9
  ) %>% 
  filter(!is.na(Total)) %>% 
  slice(c(-14,-20,-23)) %>% 
  mutate(Major = toupper(Major),
         Total = as.numeric(Total),
         Race_White_Percent = as.numeric(Race_White_Percent),
         Race_Black_Percent = as.numeric(Race_Black_Percent),
         Race_Asian_Percent = as.numeric(Race_Asian_Percent),
         Race_Hispanic_Percent = as.numeric(Race_Hispanic_Percent),
         Race_OtherMulti_Percent = as.numeric(Race_OtherMulti_Percent),
         Poverty_Percent = as.numeric(Poverty_Percent),
         Not_Poverty_Percent = as.numeric(Not_Poverty_Percent)
         )

# Count Table
sexRaceCount <- sexRaceFrequency %>%
  mutate(
    Race_White_Count = floor((Race_White_Percent / 100)*Total),
    Race_Black_Count = floor((Race_Black_Percent / 100)*Total),
    Race_Asian_Count = floor((Race_Asian_Percent / 100)*Total),
    Race_Hispanic_Count = floor((Race_Hispanic_Percent / 100)*Total),
    Race_OtherMulti_Count = floor((Race_OtherMulti_Percent / 100)*Total),
    Poverty_Count = floor((Poverty_Percent / 100)*Total),
    Not_Poverty_Count = floor((Not_Poverty_Percent / 100)*Total)
  )

# Join our two dataframes together
joinedTables <-
  inner_join(
    sexRaceCount,
    recentGrads,
    by = join_by(Major == Major)
  ) 

# Make a Summary Table
majorCategories <- joinedTables %>% 
  group_by(Major_category) %>%
  summarize(
    "Count" = sum(Total.x),
    "Average First Quartile" = mean(P25th),
    "Average Median Salary" = mean(Median),
    "Average Third Quartile" = mean(P75th),
    "Average Unemployment Rate" = mean(Unemployment_rate),
    "White" = floor(mean(Race_White_Count)),
    "Black" = floor(mean(Race_Black_Count)),
    "Asian" = floor(mean(Race_Asian_Count)),
    "Hispanic" = floor(mean(Race_Hispanic_Count)),
    "Other" = floor(mean(Race_OtherMulti_Count)),
    "Poverty Count" = floor(mean(Poverty_Count))
  )



```
## Analysis of Salaries among Major Categories

Below is an initial summary table that depicts the average salary of each quartiles 1-3, as well as the standard deviation of the median.


```{r}
#| tbl-colwidths: [75,25]

### Provides a Table on major category summary statistics

salarySummary <- recentGrads %>%
  filter(Major_category != 'Interdisciplinary') %>%
  group_by(Major_category) %>%
  summarize(
    "Count" = n(),
    "Average First Quartile" = mean(P25th),
    "Average Median Salary" = mean(Median),
    "Std. Dev. of Median Salary" = sd(Median),
    "Average Third Quartile" = mean(P75th),
  )
salarySummary %>%
  kable(
    caption = "Salary Summary Table per Major Category",
    format = "simple",
    booktabs = TRUE,
    align = "lrrrrr",
    format.args = list(big.mark = ',', digits = 2)
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "condensed"),
    font_size = 10,
  ) 
  
```

At first sight, one can see that engineering has quite a high salary for all quartiles. Even compared to the second highest earning category of major, business. We will perform a hypothesis test (two sample t test) to determine whether the average engineering median salary is significantly different from that of the business category.

  $H_0$ : $\mu_1 = \mu_2$ There is no significant difference in average median salary for engineering majors and business majors.  
  $H_a$ : $\mu_1 \ne \mu_2$ There exists a significant difference in average median salary for engineering majors and business majors.  

It is important to note, assuming engineering is sample 1 and business is sample 2:

$\bar{x}_1 = 57383$  
$s_1 = 13626$  
$n_1 = 29$  
$\bar{x}_2 = 43538$  
$s_2 = 7774$  
$n_2 = 13$  

From these numbers, the $t-value$ can be found to be:

$$t = 4.16$$
And the $p-value$ is found to be:

$$p = 0.0002$$
Because p is less than 0.05, the null hypothesis is rejected. Therefore, there is reason to believe that the median salary of engineering majors differs from that of business majors.

Now that we have this information, let's take a look at the shape of the engineering median salary data to search for outliers and any skew. See below.

```{r}
#| fig.cap = "Exploring the Shape of the Engineering Median Salary Distribution"

### Make a box plot depicting engineering salary distribution

#1 Make the necessary table, filters only engineering majors
engineerStats <- recentGrads %>%
  filter(Major_category == 'Engineering')

ggplot(engineerStats) +
  aes(x = Median, y = "" ) +
  geom_boxplot(fill = "#FF8C00") +
  labs(
    x = "Median Salary (USD)",
    y = "",
    title = "Engineering Median Salary Distribution",
  ) +
  scale_x_continuous(labels = scales::comma) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 12L),
    axis.text.x = element_text(size = 12L)
  )

```


```{r} 
### Find out info about the highest earning engineering major

#1 Find Major Name
highestEarning <- engineerStats %>%
  filter(Median == max(Median)) %>%
  select(Major) 
#2 Find Major Median Salary
highestMedSalary <- engineerStats %>%
  filter(Major == highestEarning[[1]]) %>%
  select(Median)
#3 Create table without outlier
withoutOutlier <- engineerStats %>%  
  filter(Major != 'PETROLEUM ENGINEERING')  

```
Although this plot does not seem to have a large skew one way or another, there is a noticeable outlier that lies well beyond the other median salaries. The outlier major would be `{r} highestEarning[[1]]` with a median salary of \$`{r} format(highestMedSalary[[1]], scientific = FALSE, big.mark = ",")`. The adjusted mean of the median salary without the outlier is \$`{r} format(mean(withoutOutlier$Median), scientific = FALSE, big.mark = ",")` while the standard deviation is now \$`{r} format(sd(withoutOutlier$Median), scientific = FALSE, big.mark = ",")`

``` {r}
#| fig.cap = "Median Salary per Engineering Major"
### Creates a sideways bar chart of Engineering median salaries
library(ggplot2)
#1 Select data frame to make graph from and pick which variables correspond to each axis
ggplot(engineerStats) +
  aes(x = Major, y = Median) +
  geom_col(fill = "#4081F6") +
  coord_flip() +
  theme_minimal()
```


## An Analysis of Race Frequency among Major Categories

This analysis will try to determine if there is a statistically significant bias towards people of white ethnicity in a college setting by comparing the collegiate white race percentage with the national white race percentage. We will be conducting a One Sample Proportion Z-Test to make our conclusion.

**Hypotheses**:  

  $H_0$ : $p = p_0$ There is no significant difference in the white collegiate race proportion and national (population) white race proportion.  
  $H_a$ : $p \ne p_0$ There exists a significant difference in the proportion of white people between college students and the national (population).  
  
To start, we must find the proportion of white college students. The proportion lies in the table below.

Below is a table depicting the percent frequency of each race in each major category as a portion of the sample population.

```{r}
### Make a Race/Major Category frequency table

#1 Select Necessary attributes and pivot longer
freqTablePrep <- majorCategories %>%
  select(1,7,8,9,10,11) %>%
  pivot_longer(!Major_category, names_to = "Race", values_to = "Count")
  
#2 Use uncount to have a case be one individual
byStudent <- uncount(freqTablePrep, Count)

#3 Formats frequency table and relative frequencies
raceFreq <- byStudent %>% 
  tabyl(Major_category, Race) %>%
  adorn_totals(where = "col") %>%
  arrange(desc(Total)) %>%
  adorn_totals(where = "row") %>%
  adorn_percentages(denominator = "all") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_title(
    placement = "combined",
    row_name = "Major Category",
    col_name = "Race"
  )

#4 Outputs a stylized table
raceFreq %>%
  kable(
    caption = "Frequency of Each Major Category Based on Race",
    booktabs = TRUE,
    align = "lrrrrrr"
  ) 
```

As one can tell, there is a clear disparity between the proportion of white students and the proportion of students of different races, with the total percentage of white college students being 64.14%. Since we do not yet have the population proportion, no assumptions relating to our hypotheses can be made. 

Another noteworthy finding is that 20.46% of Asian students have a major in the Computers and Mathematics Category, as opposed to just 8.13% of the whole collegiate population. This percentage can be obtained by taking the 2.38% of Asians in Computers & Mathematics and divide by the 11.63% of total Asian students in the collegiate population. This statistic tells us that Asians are more likely to have a major in Computers and Mathematics than the college population as a whole. Even still, everyone should avoid stereotyping so that all people have free choice to study whatever they choose.  

Back to our analysis, the college proportion of white students is 64.14%. Let's call this $p_0$. According to the US Census, 63.44% of the total United States population identifies as white. Let's call this $p$. According to the One Sample Proportion Z-Test:  
  
  $$z = \frac{p-p_0}{\sqrt{p_0(1-p_0)/n}}$$  
  
Where $p_0$ is the white college students proportion, $p$ is the total white US population proportion, and $n$ is the total number of college students in the sample:  
 `r sum(majorCategories['Count'])`
 
As such, our z value is calculated to be:

$$ z = -84.73 $$
Since our z value is so large (in absolute value), our p-value is most nearly 0, leading us to reject our null hypothesis. This leads us to believe there is a statistical difference in the proportion of white college students and the proportion of the US Population that is white. There are faults with this analysis, however; since the college data came from the US Census as well, such a large "sample" size would indicate a difference in the smallest of proportions. Therefore, Any insights to be made from this test are rendered insignificant. Conducting tests on a smaller scale (such as a school by school basis) will help individual school leaders find if their school is meeting diversity standards.



``` {r}
ggplot(majorCategories, aes(y = Major_category))+
  geom_boxplot(
    aes(
      xmin = `Average First Quartile`,
      xlower = `Average First Quartile`,
      xmiddle = `Average Median Salary`,
      xupper = `Average Third Quartile`,
      xmax = `Average Third Quartile`,
      fill = Major_category
    ),
    stat = "identity"
  )+
  labs(
    x = "Salary",
    y = "Major Category",
    title = "Salary Distribution by Major Category"
  )+ theme_minimal()+
  theme(
    plot.title.position = "plot",
    legend.position = "none"
  )
```

``` {r}

##Creates dot plot that shows the number of graduates from each category by race

#1 Select race attributes and pivot longer to get totals

dotPlot <- majorCategories %>%
  select(1,7,8,9,10,11) %>%
  pivot_longer(!Major_category, names_to = "Race", values_to = "Count")

#2 Creates dot plot of total gradutes by category and race

ggplot(dotPlot)+
  aes(
    x = Count,
    y = Major_category,
    color = Race#makes each race a different color
  )+
  geom_point(size = 2)+#increased size to enahce readability
  labs(
    x = "Count",
    y = "Major Category",
    title = "Total College Graduates by Major Category and Race"
  )+
  theme_minimal()+
  theme(
    plot.title.position = "plot"
  )
```
``` {r}
##Creates dot plot to show the percentages of colege graduates from each race for each major category

#1 Wrangle data to create percentage column that shows percent for each category
dotPlotPct <- dotPlot %>% 
  group_by(Race) %>% 
  mutate (Pct = Count / sum(Count))

#2 Creats dot plot of percent of total gradute from each race per major category

ggplot(dotPlotPct)+
  aes(
    x = Pct,
    y = Major_category,
    color = Race
  )+
  geom_point(size = 2)+
  labs(
    x = "Percent of Each Race",
    y = "Major Category",
    title = "Percent of College Graduates of Each Race by Major Category"
  )+
  theme_minimal()+
  theme(
    plot.title.position = "plot"
  )
```
``` {r}
## Creates dot plot that shows the percent of college graduates from each major category that are in poverty

#1 Creates dot plot
ggplot(majorCategories) +
  aes(
    x = `Poverty Count` / Count, #equation that generates percentage
    y = Major_category,
    color = Major_category
  ) +
  geom_point(size = 3) +
  labs(
    x = "Percentage in Poverty",
    y = "Major Category",
    title = "Percentage of College Graduates in Poverty by Major Category"
  )+
  theme_minimal()+
  theme(
    plot.title.position = "plot",
    legend.position = "none"
  )
```
```{r}
#Creates dot plot 
ggplot(majorCategories) +
  aes(
    x = `Poverty Count` / Count,
    y = `Average Median Salary`,
    colour = Major_category
  ) +
  geom_point(size = 3) +
  labs(
    x = "Percentage in Poverty",
    y = "Average Median Salary",
    color = "Major Category",
    title = "Total College Graduates in Poverty by Major Category"
  )+
  theme_minimal()+
  theme(
    plot.title.position = "plot"
)
```

``` {r}
ggplot(majorCategories)+
  aes(
    x = `Average Unemployment Rate`,
    y = Major_category,
    color = Major_category
  )+
  geom_point(size = 1)+
  labs(
    x = "Unemployment Rate",
    y = "Major Category",
    title = "Unemployment Rate for College Graduates by Major Category"
  )+ theme_minimal()+
  theme(
    plot.title.position = "plot",
    legend.position = "none"
  )
```

# Conclusions

Our conclusions are that

# References

Global B2B Market Research & Advisory Solutions. (2025, February 21). *United   States population by Race & Ethnicity - 2025 update*. Neilsberg.
https://www.neilsberg.com/insights/united-states-population-by-race/#:~:  
text=The%20percent%20distribution%20of%20United%20States%20population%20by,  
are%20some%20other%20race%20and%2010.71%25%20are%20multiracial. 

Tunguz, B. (2021, April 20). *College majors*. Kaggle. https://www.kaggle.com/datasets/tunguz/college-majors 

U.S. Census Bureau (2022). *American Community Survey*. Census.gov. www.census.gov/acs

```{r codeAppend, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
